Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-06-04 submission/manuscript.tex
smooth out the intro

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). The envrionment has been attributed
as an important risk factor for CRC (3). This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(4--6) and whether the bacterial community is altered (7, 8). Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colon to be
decreased in patients with carcinoma tumors (9--11). Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that is part
of our general diet (12). The most extensively studied of these SCFAs
are acetate, butyrate, and propionate (13). Overall, CRC is a growing
health problem and the microbiota through SCFAs have shown promise as a
viable treatment option.

Within model systems SCFAs, in particular butyrate, have shown positive
results in CRC treatment (14). In particular, butyrate has been shown to
inhibit cancer cell growth in \emph{in vitro} systems (15).
Additionally, supplementation with food sources that bacteria use to
create these SCFAs may also be able to confer beneficial effects. Fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while also increasing SCFA concentrations (16). Although
these model systems provide important preliminary evidence towards the
ability of SCFAs to reduce and treat tumors, the studies reporting
benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC have
found that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas and individuals
with normal colons (17). Although this would argue that increasing SCFA
concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence (18). These findings would argue
against the utility of treatments that aim to use SCFAs to reduce or
protect against tumorigenesis. Given the lack of clear evidence in human
studies of the benefit of SCFAs in CRC there is a need for more
investigation into this area.

Our current study will fill some of the current gaps in the literature
that relate to the study of SCFAs and CRC in human populations.
Specifically, it will attempt to replicate previous case/control
findings on SCFA concentrations in individuals with and without tumors.
Additionally, by increasing the adenoma tumor cohort it we will be able
to identify if the observation of a continuous gradual reduction in SCFA
concentrations occurs as tumor severity increases. It also will build
upon these observations and assess the utility of using SCFAs and
Operational Taxonomic Units (OTUs) as a potential risk stratification
tool of colorectal tumors. This study will provide important information
on the replicability of previous findings in humans by extensively
studying how SCFAs could be associated with colorectal tumors.

To accomplish this task we will directly measure the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group will be at a cross sectional point in time while
the second group will be before (pre-) and after (post-)treatment.
Additionally, we also will assess whether adding SCFA concentrations to
OTU data improves the classification of adenoma or carcinoma, use
PICRUSt and metagenomic sequencing to assess the genes involved in SCFA
synthesis, and analyze how well 16S rRNA gene sequencing can predict
SCFA concentrations. This extensive investigation will provide
additional information as to whether the taxa that are decreased in
carcionma use SCFAs to help protect against tumorigenesis.

@@

Much of the previous literature on SCFA benefit to human colorectal
cancer has been illustrated in model systems (14). Many SCFAs are
produced through the breakdown of fiber (12). A recent study in mice
found that fiber supplementation increased SCFA concentration and
decreased tumor burden (16). Additionally, SCFAs such as butyrate have
been found to inhibit tumor growth in \emph{in vitro} experiments (15).
Yet, observations in humans has been mixed. Previous case/control
studies have suggested that SCFAs are lower in individuals with
carcinoma tumors (17). However, individual studies and meta-analysis on
fiber supplementation to prevent tumor recurrence have found no benefit
(18, 19). Our results align with what has been reported in
randomized-controlled trials that SCFAs do not protect against all types
of colorectal tumors. It is possible though that there are specific
instances where SCFAs may be beneficial.

One limitation of ours and others research into the effect of SCFAs in
colorectal cancer has been that all tumors are treated as the same type.
However, there are known differences in the types of mutations that
occur (20) and treating all tumors as equal may actually hide any
benefit that could be found in specific subsets of individuals. Similar
to the idea of specific immunotherapy for specific tumors (21), SCFAs
may have beneficial effects for specific types of colorectal cancers.
Future research will need to test if this is a valid hypothesis.
Although there are limitations with the current research on SCFAs and
colorectal tumors, evidence is building that these specific metabolites
may not be protective. Yet, taxa that are associated with SCFA
production are consistently higher in normal colons than those with
carcinomas (7, 8, 22).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (7, 8,
22). Protection could be via a different pathway and other metabolites
that have not been extensively studied. Alternatively, protection may
not occur via a metabolite but instead occur through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
10)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection with chronic
inflammation replacing the role of antibiotics. Although we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that there are many exciting new avenues to explore because of
these observations.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (23, 24). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (23).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (23, 24). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (25). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (25) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (23, 26, 27). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (28).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (29) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (30). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (31). Open Reading Frames (ORFs) were
identified using Prodigal (32), counts generated using Diamond (33),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (34), and OPF alignment used the KEGG database (35).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (35,
36) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(37). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

@@

\hypertarget{ref-Schatzkin2000}{}
19. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
of effect of a low-fat, high-fiber diet on the recurrence of colorectal
adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.

\hypertarget{ref-Fearon1990}{}
20. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
21. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-normalization_Sze2017}{}
22. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-normalization_sze2017}{}
23. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
24. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
25. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Schloss2009}{}
26. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
27. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
28. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Langille2013}{}
29. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

\hypertarget{ref-Hannigan2017}{}
30. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
31. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
32. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
33. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
34. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
35. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
36. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.

\hypertarget{ref-randomforest_citation_2002}{}
37. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US (1, 2). Although there is a genetic
component to the disease, the envrionment has been attributed as an
important and large risk factor for CRC (3). This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
(4--6) and whether the bacterial community is altered (7, 8). Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colons to be
decreased in patients with carcinoma tumors (9--11). Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that are a
part of our general diet (12). The most extensively studied of these
SCFAs are acetate, butyrate, and propionate (13). Overall, CRC is a
growing health problem and the microbiota through SCFAs have shown
promise as a potential treatment option.

Specific SCFAs, like butyrate, have shown positive results for CRC
treatment within model systems (14). Butyrate has been shown to inhibit
cancer cell growth in \emph{in vitro} systems (15). Additionally,
supplementation with food sources that bacteria use to create these
SCFAs may also be able to confer beneficial effects. For example, fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while also increasing SCFA concentrations (16). Although
these model systems provide important preliminary evidence towards the
ability of SCFAs to reduce and treat tumors, the studies reporting
benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC found
that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus either patients with adenomas or
individuals with normal colons (17). Although this would argue that
increasing SCFA concentrations could be protective against
tumorigenesis, fiber supplementation in randomized controlled trials
have consistently failed to protect against tumor recurrence (18). These
findings would argue against the utility of treatments that aim to use
SCFAs to reduce or protect against tumorigenesis. Given the lack of
clear evidence in human studies of the benefit of SCFAs in CRC, there is
a need for more investigation into this area.

Our current study will fill some of the current gaps in the literature
that relate to the study of SCFAs and CRC in human populations.
Specifically, it will attempt to replicate previous case/control
findings on SCFA concentrations in individuals with and without tumors.
Additionally, by increasing the adenoma tumor cohort it we will identify
if the observation of a continuous gradual reduction in SCFA
concentrations occurs as tumor severity increases. It also will build
upon these observations and assess the utility of using SCFAs and
Operational Taxonomic Units (OTUs) as a potential risk stratification
tool of colorectal tumors (adenoma or carcinoma). This study will
provide important information on the replicability of previous findings
in humans by extensively studying how SCFAs could be associated with
colorectal tumors.

To accomplish this task we will directly measure the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group had a sample obtained at a single cross
sectional point in time while the second group had samples obtained
before (pre-) and after (post-)treatment for colorectal tumors.
Additionally, we also assessed whether adding SCFA concentrations to OTU
data improves the classification of adenoma or carcinoma using the
Random Forest algorithm (19), used PICRUSt (20) and metagenomic
sequencing to assess the genes involved in SCFA synthesis, and analyzed
how well 16S rRNA gene sequencing can predict SCFA concentrations. This
investigation provides additional information as to whether the taxa
that are decreased in carcionma use SCFAs to help protect against colon
tumorigenesis.

@@

Much of the previous literature on SCFA benefit to human colorectal
cancer has been illustrated in model systems (14). Many SCFAs are
produced through the breakdown of fiber (12). A recent study in mice
found that fiber supplementation increased SCFA concentration and
decreased tumor burden (16). Additionally, SCFAs such as butyrate have
been found to inhibit tumor growth in \emph{in vitro} experiments (15).
Yet, observations in humans has been mixed. Previous case/control
studies have suggested that SCFAs are lower in individuals with
carcinoma tumors (17). However, individual studies and meta-analysis on
fiber supplementation to prevent tumor recurrence have found no benefit
(18, 21). Our results align with what has been reported in
randomized-controlled trials that SCFAs do not protect against all types
of colorectal tumors. It is possible though that there are specific
instances where SCFAs may be beneficial.

One limitation of ours and others research into the effect of SCFAs in
colorectal cancer has been that all tumors are treated as the same type.
However, there are known differences in the types of mutations that
occur (22) and treating all tumors as equal may actually hide any
benefit that could be found in specific subsets of individuals. Similar
to the idea of specific immunotherapy for specific tumors (23), SCFAs
may have beneficial effects for specific types of colorectal cancers.
Future research will need to test if this is a valid hypothesis.
Although there are limitations with the current research on SCFAs and
colorectal tumors, evidence is building that these specific metabolites
may not be protective. Yet, taxa that are associated with SCFA
production are consistently higher in normal colons than those with
carcinomas (7, 8, 24).

The potential protection against colorectal cancers may not be from
SCFAs even though taxa associated with their production are higher in
individuals with normal colons versus patients with carcinomas (7, 8,
24). Protection could be via a different pathway and other metabolites
that have not been extensively studied. Alternatively, protection may
not occur via a metabolite but instead occur through niche exclusion of
mouth-associated microbes (e.g. \emph{Fusobacterium},
\emph{Porphyromonas}, \emph{Parvimonas}, \emph{Peptostreptococcus} (9,
10)). The idea of niche exclusion is similar to how the community
protects against \emph{Clostridium difficile} infection with chronic
inflammation replacing the role of antibiotics. Although we did not find
lower concentrations of SCFAs associated with colorectal tumors, we
think that there are many exciting new avenues to explore because of
these observations.

@@

\textbf{Study design and sampling.} The overall protocol has been
described in detail previously (25, 26). In brief, this study used fecal
samples obtained at either a single cross-sectional time point (n=490)
or from before (pre-) and after (post-) treatment for their tumor
(n=67). For patients undergoing treatment for their tumor the length of
time between their initial and follow up sample ranged from 188 - 546
days. Our use of treatment has been previously defined as encompassing
removal of a tumor with or without chemotherapy and radiation (25).
Diagnosis of tumor was made by colonoscopic examination and
histopathological review of biopsies obtained (25, 26). The University
of Michigan Institutional Review Board approved the study and informed
consent was obtained from all participants in accordance to the
guidelines set out by the Helsinki Decleration.

\textbf{Measuring specific SCFAs.} Our protocol for the measurement of
acetate, butyrate, and propionate followed a previously published
protocol (27). The following changes to this protocol included the use
of frozen fecal samples suspended in 1ml of PBS instead of fecal
suspensions in DNA Genotek OmniGut tubes, and the use of the acutal
weight of fecal samples instead of the average weight for SCFA
concentration normalizations. These changes did not affect the overall
median concentrations of these SCFAs between the two studies (see Table
1 (27) and Figure 1 in this report).

\textbf{16s rRNA gene sequencing.} The workflow and processing have been
described previously (25, 28, 29). The major differences from these
previous reports include: the use of version 1.39.5 of the mothur
software package and clustering Operational Taxonomic Units (OTUs) at
97\% similarity used the OptClust algorithm (30).

\textbf{Generating imputed metagenomes.} The use of PICRUSt version
1.1.2 with the recommended standard operating protocol (20) was used.
Briefly, the mothur shared file and metadata was converted into a biom
formated table using the biom convert function, the subsequent biom file
was processed with the normalize\_by\_copy\_number.py function, and
subsequent imputed metagenomes created using the predict\_metagenomes.py
function.

\textbf{Obtaining OPFs from metagenomes.} A subset of the
cross-sectional group (n=490) containing a total of 85 individuals
(normal n=29 normal, adenoma n=28, and carcinoma n=28) was shotgun
sequenced on an Illumina HiSeq using 125 bp paired end reads and a
previously described method (31). Briefly, the sequences were quality
filtered and sequences aligning to the human genome were removed prior
to contig assembly with MEGAHIT (32). Open Reading Frames (ORFs) were
identified using Prodigal (33), counts generated using Diamond (34),
subsequent clustering into Operational Protein Families (OPFs) used
mmseq2 (35), and OPF alignment used the KEGG database (36).

\textbf{Pulling genes involved with SCFA synthesis.} Specific genes
located near the end of the pathways involved in the synthesis of
acetate, butyrate, and propionate were analyzed for any differences
between individuals with normal colons and those with tumors. These
genes were based on pathways from KEGG as well as previous research (36,
37) and a list can be found in the supplemental material {[}Table S1{]}.

\textbf{Random Forest Models.} The model was first trained on 80\% of
the data and then tested on the held out 20\% (80/20 split) using the
Random Forest algorithm for both classification and regression models
(19). This was repeated on 100 differen 80/20 splits of the data to
generate a reasonable range for the AUC of the model. The reported AUCs,
unless otherwise specified, are for the test sets. Classification models
were built to group normal versus adenoma and normal versus carcinoma or
high versus low SCFA concentrations.

@@

\hypertarget{ref-randomforest_citation_2002}{}
19. \textbf{Liaw A}, \textbf{Wiener M}. 2002. Classification and
regression by randomForest. R News \textbf{2}:18--22.

\hypertarget{ref-Langille2013}{}
20. \textbf{Langille MGI}, \textbf{Zaneveld J}, \textbf{Caporaso JG},
\textbf{McDonald D}, \textbf{Knights D}, \textbf{Reyes JA},
\textbf{Clemente JC}, \textbf{Burkepile DE}, \textbf{Thurber RLV},
\textbf{Knight R}, \textbf{Beiko RG}, \textbf{Huttenhower C}. 2013.
Predictive functional profiling of microbial communities using 16S rRNA
marker gene sequences. Nature Biotechnology \textbf{31}:814--821.
doi:\href{https://doi.org/10.1038/nbt.2676}{10.1038/nbt.2676}.

\hypertarget{ref-Schatzkin2000}{}
21. \textbf{Schatzkin A}, \textbf{Lanza E}, \textbf{Corle D},
\textbf{Lance P}, \textbf{Iber F}, \textbf{Caan B}, \textbf{Shike M},
\textbf{Weissfeld J}, \textbf{Burt R}, \textbf{Cooper MR},
\textbf{Kikendall JW}, \textbf{Cahill J}, \textbf{Freedman L},
\textbf{Marshall J}, \textbf{Schoen RE}, \textbf{Slattery M}. 2000. Lack
of effect of a low-fat, high-fiber diet on the recurrence of colorectal
adenomas. New England Journal of Medicine \textbf{342}:1149--1155.
doi:\href{https://doi.org/10.1056/nejm200004203421601}{10.1056/nejm200004203421601}.

\hypertarget{ref-Fearon1990}{}
22. \textbf{Fearon ER}, \textbf{Vogelstein B}. 1990. A genetic model for
colorectal tumorigenesis. Cell \textbf{61}:759--767.
doi:\href{https://doi.org/10.1016/0092-8674(90)90186-i}{10.1016/0092-8674(90)90186-i}.

\hypertarget{ref-Thomas2016}{}
23. \textbf{Thomas X}, \textbf{Heiblig M}. 2016. The development of
agents targeting the BCR-ABL tyrosine kinase as philadelphia
chromosome-positive acute lymphoblastic leukemia treatment. Expert
Opinion on Drug Discovery \textbf{11}:1061--1070.
doi:\href{https://doi.org/10.1080/17460441.2016.1227318}{10.1080/17460441.2016.1227318}.

\hypertarget{ref-normalization_Sze2017}{}
24. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-normalization_sze2017}{}
25. \textbf{Sze MA}, \textbf{Baxter NT}, \textbf{Ruffin MT},
\textbf{Rogers MAM}, \textbf{Schloss PD}. 2017. Normalization of the
microbiota in patients after treatment for colonic lesions. Microbiome
\textbf{5}.
doi:\href{https://doi.org/10.1186/s40168-017-0366-3}{10.1186/s40168-017-0366-3}.

\hypertarget{ref-crc_model_baxter2016}{}
26. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-scfa_measures_venkataraman2016}{}
27. \textbf{Venkataraman A}, \textbf{Sieber JR}, \textbf{Schmidt AW},
\textbf{Waldron C}, \textbf{Theis KR}, \textbf{Schmidt TM}. 2016.
Variable responses of human microbiomes to dietary supplementation with
resistant starch. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0178-x}{10.1186/s40168-016-0178-x}.

\hypertarget{ref-Schloss2009}{}
28. \textbf{Schloss PD}, \textbf{Westcott SL}, \textbf{Ryabin T},
\textbf{Hall JR}, \textbf{Hartmann M}, \textbf{Hollister EB},
\textbf{Lesniewski RA}, \textbf{Oakley BB}, \textbf{Parks DH},
\textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres B},
\textbf{Thallinger GG}, \textbf{Horn DJV}, \textbf{Weber CF}. 2009.
Introducing mothur: Open-source, platform-independent,
community-supported software for describing and comparing microbial
communities. Applied and Environmental Microbiology
\textbf{75}:7537--7541.
doi:\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Kozich2013}{}
29. \textbf{Kozich JJ}, \textbf{Westcott SL}, \textbf{Baxter NT},
\textbf{Highlander SK}, \textbf{Schloss PD}. 2013. Development of a
dual-index sequencing strategy and curation pipeline for analyzing
amplicon sequence data on the MiSeq illumina sequencing platform.
Applied and Environmental Microbiology \textbf{79}:5112--5120.
doi:\href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-opticlust_westcott2017}{}
30. \textbf{Westcott SL}, \textbf{Schloss PD}. 2017. OptiClust, an
improved method for assigning amplicon-based sequence data to
operational taxonomic units. mSphere \textbf{2}:e00073--17.
doi:\href{https://doi.org/10.1128/mspheredirect.00073-17}{10.1128/mspheredirect.00073-17}.

\hypertarget{ref-Hannigan2017}{}
31. \textbf{Hannigan GD}, \textbf{Duhaime MB}, \textbf{Ruffin MT},
\textbf{Koumpouras CC}, \textbf{Schloss PD}. 2017. Diagnostic potential
\& the interactive dynamics of the colorectal cancer virome.
doi:\href{https://doi.org/10.1101/152868}{10.1101/152868}.

\hypertarget{ref-Li2015}{}
32. \textbf{Li D}, \textbf{Liu C-M}, \textbf{Luo R}, \textbf{Sadakane
K}, \textbf{Lam T-W}. 2015. MEGAHIT: An ultra-fast single-node solution
for large and complex metagenomics assembly via succinct de bruijn
graph. Bioinformatics \textbf{31}:1674--1676.
doi:\href{https://doi.org/10.1093/bioinformatics/btv033}{10.1093/bioinformatics/btv033}.

\hypertarget{ref-Hyatt2010}{}
33. \textbf{Hyatt D}, \textbf{Chen G-L}, \textbf{LoCascio PF},
\textbf{Land ML}, \textbf{Larimer FW}, \textbf{Hauser LJ}. 2010.
Prodigal: Prokaryotic gene recognition and translation initiation site
identification. BMC Bioinformatics \textbf{11}:119.
doi:\href{https://doi.org/10.1186/1471-2105-11-119}{10.1186/1471-2105-11-119}.

\hypertarget{ref-Buchfink2014}{}
34. \textbf{Buchfink B}, \textbf{Xie C}, \textbf{Huson DH}. 2014. Fast
and sensitive protein alignment using DIAMOND. Nature Methods
\textbf{12}:59--60.
doi:\href{https://doi.org/10.1038/nmeth.3176}{10.1038/nmeth.3176}.

\hypertarget{ref-Steinegger2017}{}
35. \textbf{Steinegger M}, \textbf{Söding J}. 2017. MMseqs2 enables
sensitive protein sequence searching for the analysis of massive data
sets. Nature Biotechnology.
doi:\href{https://doi.org/10.1038/nbt.3988}{10.1038/nbt.3988}.

\hypertarget{ref-Kanehisa2015}{}
36. \textbf{Kanehisa M}, \textbf{Sato Y}, \textbf{Kawashima M},
\textbf{Furumichi M}, \textbf{Tanabe M}. 2015. KEGG as a reference
resource for gene and protein annotation. Nucleic Acids Research
\textbf{44}:D457--D462.
doi:\href{https://doi.org/10.1093/nar/gkv1070}{10.1093/nar/gkv1070}.

\hypertarget{ref-scfa_baxter2014}{}
37. \textbf{Baxter NT}, \textbf{Zackular JP}, \textbf{Chen GY},
\textbf{Schloss PD}. 2014. Structure of the gut microbiome following
colonization with human feces determines colonic tumor burden.
Microbiome \textbf{2}:20.
doi:\href{https://doi.org/10.1186/2049-2618-2-20}{10.1186/2049-2618-2-20}.
